\pkg{regsem}: Regularized Structural Equation Modeling

### 2017-03-22 draft1.tex
revision

\Abstract{
The \pkg{regsem} package in \proglang{R}, an implementation of
regularized structural equation modeling (RegSEM; Jacobucci, Grimm, and
McArdle 2016), was recently developed with the goal incorporating
various forms of penalized likelihood estimation with a broad array of
structural equations models. The forms of regularization include both
the \textit{ridge} (Hoerl and Kennard 1970) and the least absolute
shrinkage and selection operator (\textit{lasso}; Tibshirani 1996),
along with sparser extensions. RegSEM is particularly useful for
structural equation models that have a small parameter to sample size
ratio, as the addition of penalties can reduce the complexity, thus
reducing the bias of the parameter estimates. The paper details the
algorithmic details and an overview of the use of \pkg{regsem} with the
application of both factor analysis and latent growth curve models.
}

@@

\setlength\parindent{24pt} \setlength{\parskip}{0.5em}

@@

The desire for simplicity in model structure comes by many names,
including simple structure (Thurstone 1935), variable complexity (Browne
2001), parsimony (Raykov and Marcoulides 1999; Marsh and Hau 1996),
``sparse loadings'' in the context of principal components analysis
(Zou, Hastie, and Tibshirani 2006), and lastly, ``sparsistency'',
denoting that all parameters in a sparse model that are zero are
correctly estimated as zero with probability tending to one (Lam and Fan
2009). The goal is to accurately and efficiently estimate a model that
is parsimonious in allowing users to easily interpret the model's
representation of reality. In the context of latent variables, reducing
the complexity of models can come in many forms: selecting among
multiple predictors of a latent variable, simplifying factor structure
by removing cross-loadings, determining whether the addition of
nonlinear terms are necessary in longitudinal models, and many others.
As a simple running example, Figure 1 depicts a linear latent growth
curve model (e.g. Meredith and Tisak 1990) with four time points and ten
predictors for a simulated dataset.

@@

As an example, a researcher may want to test this model, but may only
have a relatively small sample size (e.g.~80). However, there are 29
estimated parameters in this model, resulting in a estimated parameter
to sample size ratio as far below even the most liberal recommendations
(e.g.~10:1 parameters to sample size; Kline 2015). In lieu of finding
additional respondents, reducing the number of parameters estimated is
one effective strategy for reducing bias. Specifically, the 20 estimated
regressions from \textit{c1-c10} could be reduced to a number that makes
the ratio of the parameters estimated to sample size more reasonable. To
explore this further, the next section provides an overview of
regularization, and how different forms can be used to perform variable
selection across a broad range of models.

@@

Although a whole host of methods exist to perform variable selection,
the use of regularization has seen a wide array of application in the
context of regression, and more recently, in areas such as graphical
modeling, as well as a host of others. The two most common procedures
for regularization in regression are the \textit{ridge} (Hoerl and
Kennard 1970) and the least absolute shrinkage and selection operator
(\textit{lasso}; Tibshirani 1996); however, there are various
alternative forms that can be seen as subsets or generalizations of
these two procedures. Given an outcome vector \textit{y} and predictor
matrix \(X \in {R}^{n \times p}\) , the ridge estimates are defined as

\[\tag{1}
\hat{\beta}^{ridge}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p} \beta_{j}^{2}\Big\},
\] \noindent
where \(\beta_{0}\) is the intercept, \(\beta_{j}\) is the coefficient
for \(x_{j}\), and \(\lambda\) is the penalty that controls the amount
of shrinkage. Note that when \(\lambda = 0\), Equation 3 reduces to
ordinary least squares regression. As \(\lambda\) is increased, the
\(\beta\) parameters are shrunken towards zero. The lasso estimates are
defined as

@@

\noindent
In lasso regression, the \(l_{1}\)-norm is used, instead of
\(l_{2}\)-norm as in ridge, which also shrinks the \(\beta\) parameters,
but additionally drives the parameters all the way to zero, thus
performing a form of subset selection.

@@

\noindent
where \(\Sigma\) is the model implied covariance matrix, \(C\) is the
observed covariance matrix, and \(p\) is the number of estimated
parameters. RegSEM builds in an additional element to penalize certain
model parameters yielding

@@

\noindent
where \(\lambda\) is the regularization parameter and takes on a value
between zero and infinity. When \(\lambda\) is zero, MLE is performed,
and when \(\lambda\) is infinity, all penalized parameters are shrunk to
zero. \(P(\cdot)\) is a general function for summing the values of one
or more of the model's parameter matrices. The two common forms of
\(P(\cdot)\) include both the lasso (\(\| \cdot \|_{1}\)), which
penalizes the sum of the absolute values of the parameters, and ridge
(\(\| \cdot \|_{2}\)), which penalizes the sum of the squared values of
the parameters.

@@

\[\tag{5}
F_{lasso} = F_{ML} + \lambda * \left\|  \begin{matrix}  
c1 \xrightarrow[]{} i\\
c2\xrightarrow[]{}i\\
\vdots \\
c10\xrightarrow[]{}i\\
c1\xrightarrow[]{}s\\
\vdots \\
c10\xrightarrow[]{}s\\
\end{matrix}  \right\|_{1}
\] \noindent
Although the fit of the model is easily calculated given a set of
parameter estimates, traditional optimization procedures for SEM cannot
be used given the non-differentiable nature of lasso penalties, and as
detailed later, sparse extensions.

@@

\begin{equation}
S_{s^{t} \lambda}(z_{j}) = sign(z_{j})(|z_{j}|-s^{t} \lambda)_{+}
\end{equation}

\noindent
where \((x)_{+}\) is shorthand for max(x,0) and \(s^{t}\) is the step
size. Henceforth, \(\lambda\) is assumed to encompass both the penalty
and the step size \(s^{t}\). This procedure is only used to update
parameters that are subject to penalty. Non-penalized parameters are
updated only using step 1 from above.

In contrast to only updating one parameter at a time in a
coordinate-wise fashion, for RegSEM the optimization steps can be
divided by both the \textit{A} and \textit{S} matrices. This block-wise
gradient descent manifests itself as:

\noindent

@@

\noindent
where step 13 is the calculation of the implied covariance matrix using
RAM matrices and step 14 calculates the fit of the model with penalizing
parameters in the \textit{A} matrix. Note that parameters from the
\textit{S} matrix can also be penalized, however, this is much less
common. This algorithm can be described as first order proximal block
coordinate descent. For some SEM models, using block updates has been
found to work better than standard gradient descent with the same soft
thresholding of penalized parameters.

@@

\noindent
When \(\alpha\) is zero, ridge is performed, and conversely when
\(\alpha\) is 1, lasso regularization is performed. This method
harnesses the benefits of both methods, particularly when variable
selection is warranted (lasso), but there may be collinearity between
the variables (ridge).

@@

In using lasso penalties, difficulties emerge when the scale of
variables differ dramatically. By only using one value of \(\lambda\),
this can add appreciable bias to the resulting estimates (e.g. Fan and
Li 2001). One method proposed for overcoming this limitation is the
adaptive lasso (Zou 2006). Instead of penalizing parameters directly,
each parameter is scaled by the un-penalized estimated (MLE parameter
estimates in SEM). The adaptive lasso results in: \[
F_{alasso} = F_{ML} + \lambda \| \theta_{ML}^{-1} * \theta_{pen} \|_{1}
\] \noindent
with, following the same form for the lasso, the soft-thresholding
update is:

\[
S(\theta_{j})= sign(\theta_{j})(|\theta_{j}|-\frac{\lambda}{2|\theta_{j}|})_{+}
\] \noindent
In this, larger penalties are given for non-significant (smaller)
parameters, limiting the bias in estimating larger, significant
parameters. Note that one limitation of this approach for SEM models is
that the model needs to be estimable with MLE. Particularly for models
with large numbers of variables, in relation to sample size, this may
not be possible.

@@

\noindent
with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
S(\theta_{j},\lambda),&  |\theta_{j}| \geq 2\lambda\\
\frac{\gamma-1}{\gamma-2}S(\theta_{j},\frac{\lambda\gamma}{\gamma-1}),              & 2\lambda < |\theta_{j}|\leq\alpha\lambda\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

\noindent
for \(\gamma > 2\). As the the penalty in equation 11 is non-convex (as
is the MCP), this makes the computation more difficult. However, in the
context of SEM this can be seen as less problematic, as equation 3 is
also non-convex.

@@

Zhang (2010) proposed the minimax concave penalty:

@@

\noindent
with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
\frac{\gamma}{\gamma-1}S(\theta_{j},\lambda),&  |\theta_{j}| \leq \lambda\gamma\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

\noindent
for \(\gamma > 0\). As seen in Figure 2, this results in similar amount
of shrinkage for smaller estimates in comparison to the SCAD, however,
less for larger estimates. For both the SCAD and MCP, both the
\(\gamma\) and \(\lambda\) parameters are used as hyper-parameters. This
involves testing models over a two-dimensional array of parameters,
however, \(\gamma\) can be fixed to 3.7 per Fan and Li (2001).

@@

Syntax for using the \pkg{regsem} is based on the \pkg{lavaan} package
(Rosseel 2012) for structural equation models. \pkg{lavaan} is a general
SEM software program that can fit a wide array of models with various
estimation methods. To use \pkg{regsem}, the user has to first fit the
model in \pkg{lavaan}. Note that particularly in cases that the number
of variables is larger than the sample size, the model in lavaan does
not need to converge, let alone run. In this case, the
\code{do.fit=FALSE} argument in lavaan can be used. \#\#\#maybe move
previous sentences\#\#\# Below is the code for a one factor model fit in
lavaan with the Holzinger and Swineford (1939) dataset.

@@

\begin{CodeChunk}
\begin{CodeInput}
head(round(out.reg[[1]],2),5)
\end{CodeInput}
\begin{CodeOutput}
     f1 -> x1 f1 -> x2 f1 -> x3 f1 -> x4 f1 -> x5 f1 -> x6 f1 -> x7
[1,]     0.51     0.26     0.25     0.98     1.08     0.92     0.20
[2,]     0.41     0.17     0.17     0.89     0.98     0.83     0.12
[3,]     0.33     0.10     0.10     0.83     0.91     0.77     0.06
[4,]     0.27     0.04     0.04     0.78     0.87     0.72     0.01
[5,]     0.21     0.00     0.00     0.75     0.83     0.69     0.00
     f1 -> x8 f1 -> x9 x1 ~~ x1 x2 ~~ x2 x3 ~~ x3 x4 ~~ x4 x5 ~~ x5
[1,]     0.20     0.31     1.10     1.31     1.21     0.38     0.49
[2,]     0.13     0.23     1.11     1.32     1.22     0.38     0.48
[3,]     0.08     0.17     1.13     1.33     1.23     0.38     0.47
[4,]     0.03     0.12     1.15     1.35     1.25     0.38     0.46
[5,]     0.00     0.07     1.18     1.36     1.26     0.39     0.46
     x6 ~~ x6 x7 ~~ x7 x8 ~~ x8 x9 ~~ x9
[1,]     0.36     1.15     0.98     0.92
[2,]     0.36     1.15     0.99     0.93
[3,]     0.36     1.16     1.00     0.94
[4,]     0.37     1.17     1.01     0.96
[5,]     0.38     1.18     1.02     0.97
\end{CodeOutput}
\end{CodeChunk}

\noindent
while \code{out.reg[[2]]} looks like:

\begin{CodeChunk}
\begin{CodeInput}
round(out.reg[[2]],2)
\end{CodeInput}
\begin{CodeOutput}
      lambda conv rmsea     BIC
 [1,]   0.00    0  0.19 7805.18
 [2,]   0.05    0  0.19 7815.39
 [3,]   0.10    0  0.20 7840.59
 [4,]   0.15    0  0.21 7875.84
 [5,]   0.20    0  0.20 7886.15
 [6,]   0.25    0  0.21 7911.82
 [7,]   0.30    0  0.21 7934.45
 [8,]   0.35    0  0.22 7960.67
 [9,]   0.40    0  0.23 7991.03
[10,]   0.45    0  0.23 8009.42
[11,]   0.50    0  0.23 8034.81
[12,]   0.55    0  0.24 8069.49
[13,]   0.60    0  0.29 8360.54
[14,]   0.65    0  0.29 8360.53
[15,]   0.70    0  0.29 8360.52
\end{CodeOutput}
\end{CodeChunk}

In this, the user can examine the penalty (lambda), whether the model
converged (``conv''=0), and the fit of each model. By default, two fit
indices are output, both the root mean square error of approximation
(RMSEA; Steiger and Lind 1980), and the Bayesian information criteria
(BIC; Schwarz 1978). Both the RMSEA and BIC take into account the
degrees of freedom of the model, an important point for model selection
in the presence of lasso penalties (and other penalties that set
parameters to zero). Based on work done on degrees of freedom for lasso
regression, as Zou, Hastie, and Tibshirani (2007) proved that the number
of nonzero coefficients is an unbiased estimate of the degrees of
freedom for regression. As the penalty increases, select parameter are
set to zero, thus the degrees of freedom increases, which for fit
indices that include the degrees of freedom in the calculation, means
that although the likelihood of the model may only get worse (increase),
both the RMSEA and BIC can improve (decrease) due to the impact of
increased degrees of freedom.

@@

\noindent
These same fit measures can be accessed through \code{cv_regsem},
through changing the defaults with the \code{fit.ret=c("rmsea","BIC")}
argument. Finally, instead of assessing these fit indices on the same
sample that the models were run on, a holdout dataset could be used.
This can be done two ways: either with
\code{cv_regsem(...,fit.ret2="test")} or with
\code{fit_indices(model,CV=TRUE,CovMat=)} and specifying the name of the
holdout covariance matrix.

@@

To compare different types of penalties in \pkg{regsem}, it requires a
different specification of the \code{type} argument. The options
currently include maximum likelihood (\code{"none"}), ridge
(\code{"ridge"}), lasso (\code{"lasso"}), adaptive lasso
(\code{"alasso"}), elastic net (\code{"enet"}), SCAD (\code{"scad"}),
and MCP (\code{"mcp"}). For the elastic net, there is an additional
hyperparameter, \(\alpha\) that controls the tradeoff between ridge and
lasso penalties. This is specified as \code{alpha=} , which has a
default of 0.5. Additionally, both the SCAD and MCP have the additional
hyper parameter of \(\gamma\), which is specified as \code{gamma=} and
defaults to 3.7 per Fan and Li (2001).

@@

To compare the performance of each penalization method further,
particularly in the presence of small number of parameters to sample
size ratios, a small scale simulation study was conducted. The same
model and effects was kept, but the sample size was varied to include
80, 200, and 1000 to demonstrate how maximum likelihood estimation of
effects improves as sample size increases, while each of the
regularization methods performs well regardless of sample size. Each run
was replicated 200 times. For each regularization method, the BIC was
used to choose a final model among the 40 penalty vales. The results are
displayed in Table 2.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & N & ML & lasso & alasso & SCAD & MCP \\ 
  \hline
 & 80.00 & 0.08 & 0.08 & \textbf{0.04} & 0.05 & \textbf{0.20} \\ 
  False Positives & 200.00 & 0.06 & 0.05 & \textbf{0.02} & 0.03 & 0.06 \\ 
   & 1000.00 & 0.05 & 0.08 & \textbf{0.01} & \textbf{0.01} & 0.02 \\ 
   & 80.00 & 0.33 & \textbf{0.31} & 0.35 & 0.35 & 0.31 \\ 
  False Negatives & 200.00 & \textbf{0.19} & \textbf{0.19} & 0.23 & 0.26 & 0.27 \\ 
   & 1000.00 & \textbf{0.00} & \textbf{0.00} & \textbf{0.00} & 0.01 & 0.03 \\ 
   \hline
\end{tabular}
    \caption{Results from the simulation using the model in Figure 1. Each condition was replicated 200 times. False positives represent concluding that the simulated regressions of zero were concluded as nonzero. False negatives are concluding that either the simulated regression values of 1 or 0.2 are in fact zero. Bolded values represent the smallest error per condition.}
\end{table}

For false positive, the adaptive lasso demonstrated the best
performance, where the performance of maximum likelihood estimation
leveled off at the 0.05 level at a sample size of 1000 as expected. For
false negatives, lasso penalties demonstrated similar results to maximum
likelihood. This was expected given the tendency of the lasso to
under-penalize small coefficients in comparison to the other
regularization methods. The adaptive lasso and SCAD demonstrated
slightly worse results, however, outside of the MCP, each method made
either zero or near zero errors at a sample size of 1000. The poor
performance of the MCP may be in part due to fixing the \(\gamma\)
penalty to 3.7. Varying this parameter may improve the performance of
the method. In summary, the regularization methods demonstrated an
improvement over maximum likelihood, particularly at small samples, for
a model that had a large number of estimated parameters.

@@

Broadly speaking, there is a growing amount of research into the
integration between data mining methods and latent variable models.
Specifically, beyond RegSEM, this has taken the form of item response
theory and regularization (Sun et al. 2016), pairing both structural
equation models with decision trees (Brandmaier et al. 2013),
exploratory psychological network analaysis (e.g. Epskamp, Rhemtulla,
and Borsboom 2016), along with many others. The amount of pairing
between methods that have generally been housed in separate camps will
only increase into the future. This type of research will be facilitated
by the general upsurge in the creation of open source software that
gives users a general framework to test models. This was the motivation
behind creating the \pkg{regsem} package in that users can estimate
models ranging from simple factor analysis models, to latent
longitudinal models with few to many time points, and finally to models
with a large number of latent and observed variables. The use of
regularization allows for the estimation of much larger structural
equation models than before. Specifically, given that latent variables,
as discussed in this article, are more commonly used in the social and
behavioral sciences, sample sizes are generally not large. To estimate
large models with small sample sizes invites increasing amounts of bias,
and as demonstrated with the simulated data in this paper.
Regularization can be used to reduce the complexity of the model, thus
decreasing both the bias and variance.

>>>>

\Abstract{
The \pkg{regsem} package in \proglang{R}, an implementation of
regularized structural equation modeling (RegSEM; Jacobucci, Grimm, and
McArdle 2016), was recently developed with the goal of incorporating
various forms of penalized likelihood estimation with a broad array of
structural equations models. The forms of regularization include both
the \textit{ridge} (Hoerl and Kennard 1970) and the least absolute
shrinkage and selection operator (\textit{lasso}; Tibshirani 1996),
along with sparser extensions. RegSEM is particularly useful for
structural equation models that have a small parameter to sample size
ratio, as the addition of penalties can reduce the complexity, thus
reducing the bias of the parameter estimates. The paper covers the
algorithmic details and an overview of the use of \pkg{regsem} with the
application of both factor analysis and latent growth curve models.
}

@@

@@

The desire for simplicity in model structure comes by many names,
including simple structure (Thurstone 1935), variable complexity (Browne
2001), parsimony (Raykov and Marcoulides 1999; Marsh and Hau 1996),
``sparse loadings'' in the context of principal components analysis
(Zou, Hastie, and Tibshirani 2006), and lastly, ``sparsistency'',
denoting that all parameters in a sparse model that are zero are
correctly estimated as zero with probability tending to one (Lam and Fan
2009). The goal is to accurately and efficiently estimate a model that
is parsimonious in allowing users to easily interpret the model's
representation of reality. In the context of latent variables, reducing
the complexity of models can come in many forms: selecting among
multiple predictors of a latent variable, simplifying factor structure
by removing cross-loadings, determining whether the addition of
nonlinear terms are necessary in longitudinal models, and many others.

@@

As a simple running example, Figure 1 depicts a linear latent growth
curve model (e.g. Meredith and Tisak 1990) with four time points and ten
predictors for a simulated dataset. In this, a researcher may want to
test this model, but may only have a relatively small sample size
(e.g.~80). There are 29 estimated parameters in this model, resulting in
a estimated parameter to sample size ratio as far below even the most
liberal recommendations (e.g.~10:1 parameters to sample size; Kline
2015). In lieu of finding additional respondents, reducing the number of
parameters estimated is one effective strategy for reducing bias.
Specifically, the 20 estimated regressions from \textit{c1-c10} could be
reduced to a number that makes the ratio of the parameters estimated to
sample size more reasonable. To explore this further, the next section
provides an overview of regularization, and how different forms can be
used to perform variable selection across a broad range of models.

@@

Although a whole host of methods exist to perform variable selection,
the use of regularization has seen a wide array of application in the
context of regression, and more recently, in areas such as graphical
modeling, as well as a host of others. The two most common procedures
for regularization in regression are the \textit{ridge} (Hoerl and
Kennard 1970) and the least absolute shrinkage and selection operator
(\textit{lasso}; Tibshirani 1996); however, there are various
alternative forms that can be seen as subsets or generalizations of
these two procedures. Given an outcome vector \textit{y} and predictor
matrix \(X \in {R}^{n \times p}\) , ridge estimates are defined as

\[\tag{1}
\hat{\beta}^{ridge}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p} \beta_{j}^{2}\Big\},
\]

where \(\beta_{0}\) is the intercept, \(\beta_{j}\) is the coefficient
for \(x_{j}\), and \(\lambda\) is the penalty that controls the amount
of shrinkage. Note that when \(\lambda = 0\), Equation 3 reduces to
ordinary least squares regression. As \(\lambda\) is increased, the
\(\beta\) parameters are shrunken towards zero. The lasso estimates are
defined as

@@

In lasso regression, the \(l_{1}\)-norm is used, instead of
\(l_{2}\)-norm as in ridge, which also shrinks the \(\beta\) parameters,
but additionally drives the parameters all the way to zero, thus
performing a form of subset selection.

@@

where \(\Sigma\) is the model implied covariance matrix, \(C\) is the
observed covariance matrix, and \(p\) is the number of estimated
parameters. RegSEM builds in an additional element to penalize certain
model parameters yielding

@@

where \(\lambda\) is the regularization parameter and takes on a value
between zero and infinity. When \(\lambda\) is zero, MLE is performed,
and when \(\lambda\) is infinity, all penalized parameters are shrunk to
zero. \(P(\cdot)\) is a general function for summing the values of one
or more of the model's parameter matrices. Two common forms of
\(P(\cdot)\) include both the lasso (\(\| \cdot \|_{1}\)), which
penalizes the sum of the absolute values of the parameters, and ridge
(\(\| \cdot \|_{2}\)), which penalizes the sum of the squared values of
the parameters.

@@

\[\tag{5}
F_{lasso} = F_{ML} + \lambda * \left\|  \begin{matrix}  
c1 \xrightarrow[]{} i\\
c2\xrightarrow[]{}i\\
\vdots \\
c10\xrightarrow[]{}i\\
c1\xrightarrow[]{}s\\
\vdots \\
c10\xrightarrow[]{}s\\
\end{matrix}  \right\|_{1}
\]

Although the fit of the model is easily calculated given a set of
parameter estimates, traditional optimization procedures for SEM cannot
be used given the non-differentiable nature of lasso penalties, and as
detailed later, sparse extensions.

@@

\begin{equation}
S_{s^{t} \lambda}(z_{j}) = sign(\theta_{j})(|\theta_{j}|-s^{t} \lambda)_{+}
\end{equation}

where \((x)_{+}\) is shorthand for max(x,0) and \(s^{t}\) is the step
size. Henceforth, \(\lambda\) is assumed to encompass both the penalty
and the step size \(s^{t}\). This procedure is only used to update
parameters that are subject to penalty. Non-penalized parameters are
updated only using step 1 from above.

In contrast to only updating one parameter at a time in a
coordinate-wise fashion, for RegSEM the optimization steps can be
divided by both the \textit{A} and \textit{S} matrices (see the
Implementation section for more description on the RAM matrices). This
block-wise gradient descent manifests itself as:

@@

where step 13 is the calculation of the implied covariance matrix using
RAM matrices and step 14 calculates the fit of the model with penalizing
parameters in the \textit{A} matrix. Note that parameters from the
\textit{S} matrix can also be penalized, however, this is much less
common. This algorithm can be described as first order proximal block
coordinate descent. For some SEM models, using block updates has been
found to work better than standard gradient descent with the same soft
thresholding of penalized parameters.

@@

When \(\alpha\) is zero, ridge is performed, and conversely when
\(\alpha\) is 1, lasso regularization is performed. This method
harnesses the benefits of both methods, particularly when variable
selection is warranted (lasso), but there may be collinearity between
the variables (ridge).

@@

In using lasso penalties, difficulties emerge when the scale of
variables differ dramatically. By only using one value of \(\lambda\),
this can add appreciable bias to the resulting estimates (e.g. Fan and
Li 2001). One method proposed for overcoming this limitation is the
adaptive lasso (Zou 2006). Instead of penalizing parameters directly,
each parameter is scaled by the un-penalized estimated (MLE parameter
estimates in SEM). The adaptive lasso results in: \[
F_{alasso} = F_{ML} + \lambda \| \theta_{ML}^{-1} * \theta_{pen} \|_{1}
\]

with, following the same form for the lasso, the soft-thresholding
update is:

\[
S(\theta_{j})= sign(\theta_{j})(|\theta_{j}|-\frac{\lambda}{2|\theta_{j}|})_{+}
\]

In this, larger penalties are given for non-significant (smaller)
parameters, limiting the bias in estimating larger, significant
parameters. Note that one limitation of this approach for SEM models is
that the model needs to be estimable with MLE. Particularly for models
with large numbers of variables, in relation to sample size, this may
not be possible.

@@

with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
S(\theta_{j},\lambda),&  |\theta_{j}| \geq 2\lambda\\
\frac{\gamma-1}{\gamma-2}S(\theta_{j},\frac{\lambda\gamma}{\gamma-1}),              & 2\lambda < |\theta_{j}|\leq\alpha\lambda\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

for \(\gamma > 2\). As the the penalty in equation 11 is non-convex (as
is the MCP), this makes the computation more difficult. However, in the
context of SEM this can be seen as less problematic, as equation 3 is
also non-convex.

@@

A final regularization method developed to produce sparser penalties
than the lasso was the minimax concave penalty (MCP, Zhang 2010):

@@

with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
\frac{\gamma}{\gamma-1}S(\theta_{j},\lambda),&  |\theta_{j}| \leq \lambda\gamma\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

for \(\gamma > 0\). As seen in Figure 2, this results in similar amount
of shrinkage for smaller estimates in comparison to the SCAD, however,
less for larger estimates. For both the SCAD and MCP, both the
\(\gamma\) and \(\lambda\) parameters are used as hyper-parameters. This
involves testing models over a two-dimensional array of parameters,
however, in \pkg{regsem}, \(\gamma\) is by default fixed to 3.7 per Fan
and Li (2001).

@@

Syntax for using the \pkg{regsem} is based on the \pkg{lavaan} package
(Rosseel 2012) for structural equation models. \pkg{lavaan} is a general
SEM software program that can fit a wide array of models with various
estimation methods. To use \pkg{regsem}, the user has to first fit the
model in \pkg{lavaan}. Note that particularly in cases that the number
of variables is larger than the sample size, the model in lavaan does
not need to converge, let alone run. In this case, the
\code{do.fit=FALSE} argument in lavaan can be used. As a canonical
example, below is the code for a confirmatory factor analysis model with
one latent factor and nince indicators from the Holzinger and Swineford
(1939) dataset.

@@

\begin{CodeChunk}
\begin{CodeInput}
head(round(out.reg[[1]],2),5)
\end{CodeInput}
\begin{CodeOutput}
     f1 -> x1 f1 -> x2 f1 -> x3 f1 -> x4 f1 -> x5 f1 -> x6 f1 -> x7
[1,]     0.51     0.26     0.25     0.98     1.08     0.92     0.20
[2,]     0.41     0.17     0.17     0.89     0.98     0.83     0.12
[3,]     0.33     0.10     0.10     0.83     0.91     0.77     0.06
[4,]     0.27     0.04     0.04     0.78     0.87     0.72     0.01
[5,]     0.21     0.00     0.00     0.75     0.83     0.69     0.00
     f1 -> x8 f1 -> x9 x1 ~~ x1 x2 ~~ x2 x3 ~~ x3 x4 ~~ x4 x5 ~~ x5
[1,]     0.20     0.31     1.10     1.31     1.21     0.38     0.49
[2,]     0.13     0.23     1.11     1.32     1.22     0.38     0.48
[3,]     0.08     0.17     1.13     1.33     1.23     0.38     0.47
[4,]     0.03     0.12     1.15     1.35     1.25     0.38     0.46
[5,]     0.00     0.07     1.18     1.37     1.26     0.39     0.46
     x6 ~~ x6 x7 ~~ x7 x8 ~~ x8 x9 ~~ x9
[1,]     0.36     1.15     0.98     0.92
[2,]     0.36     1.15     0.99     0.93
[3,]     0.36     1.16     1.00     0.94
[4,]     0.37     1.17     1.01     0.96
[5,]     0.38     1.18     1.02     0.97
\end{CodeOutput}
\end{CodeChunk}

while \code{out.reg[[2]]} contains information pertaining to the fit of
each model:

\begin{CodeChunk}
\begin{CodeInput}
round(out.reg[[2]],2)
\end{CodeInput}
\begin{CodeOutput}
      lambda conv rmsea     BIC
 [1,]   0.00    0  0.19 7805.18
 [2,]   0.05    0  0.19 7815.39
 [3,]   0.10    0  0.20 7840.40
 [4,]   0.15    0  0.21 7875.41
 [5,]   0.20    0  0.20 7885.68
 [6,]   0.25    0  0.21 7910.92
 [7,]   0.30    0  0.21 7933.99
 [8,]   0.35    0  0.22 7960.28
 [9,]   0.40    0  0.23 7990.80
[10,]   0.45    0  0.23 8007.95
[11,]   0.50    0  0.23 8034.00
[12,]   0.55    0  0.24 8069.15
[13,]   0.60    0  0.29 8360.53
[14,]   0.65    0  0.29 8360.53
[15,]   0.70    0  0.29 8360.51
\end{CodeOutput}
\end{CodeChunk}

In this, the user can examine the penalty (lambda), whether the model
converged (``conv''=0), and the fit of each model. By default, two fit
indices are output, both the root mean square error of approximation
(RMSEA; Steiger and Lind 1980), and the Bayesian information criteria
(BIC; Schwarz 1978). Both the RMSEA and BIC take into account the
degrees of freedom of the model, an important point for model selection
in the presence of lasso penalties (and other penalties that set
parameters to zero). Zou, Hastie, and Tibshirani (2007) proved that the
number of nonzero coefficients is an unbiased estimate of the degrees of
freedom for regression. As the penalty increases, select parameters are
set to zero, thus increasing the degrees of freedom, which for fit
indices that include the degrees of freedom in the calculation, means
that although the likelihood of the model may only get worse (increase),
both the RMSEA and BIC can improve (decrease).

@@

These same fit measures can be accessed through \code{cv_regsem} through
changing the defaults with the \code{fit.ret=c("rmsea","BIC")} argument.
Finally, instead of assessing these fit indices on the same sample that
the models were run on, a holdout dataset could be used. This can be
done two ways: either with \code{cv_regsem(...,fit.ret2="test")} or with
\code{fit_indices(model,CV=TRUE,CovMat=)} and specifying the name of the
holdout covariance matrix.

@@

Comparing different types of penalties in \pkg{regsem} requires a
different specification of the \code{type} argument. The options
currently include maximum likelihood (\code{"none"}), ridge
(\code{"ridge"}), lasso (\code{"lasso"}), adaptive lasso
(\code{"alasso"}), elastic net (\code{"enet"}), SCAD (\code{"scad"}),
and MCP (\code{"mcp"}). For the elastic net, there is an additional
hyperparameter, \(\alpha\) that controls the tradeoff between ridge and
lasso penalties. This is specified as \code{alpha=} , which has a
default of 0.5. Additionally, both the SCAD and MCP have the additional
hyper parameter of \(\gamma\), which is specified as \code{gamma=} and
defaults to 3.7 per Fan and Li (2001).

@@

To compare the performance of each penalization method further,
particularly in the presence of a small parameter to sample size ratio,
a small simulation study was conducted. The same model and effects was
kept, but the sample size was varied to include 80, 200, and 1000 to
demonstrate how maximum likelihood estimation of effects improves as
sample size increases, while each of the regularization methods performs
well regardless of sample size. Each run was replicated 200 times. For
each regularization method, the BIC was used to choose a final model
among the 40 penalty vales. The results are displayed in Table 2.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & N & ML & lasso & alasso & SCAD & MCP \\ 
  \hline
 & 80.00 & 0.08 & 0.08 & \textbf{0.04} & 0.05 & 0.20 \\ 
  False Positives & 200.00 & 0.06 & 0.05 & \textbf{0.02} & 0.03 & 0.06 \\ 
   & 1000.00 & 0.05 & 0.08 & \textbf{0.01} & \textbf{0.01} & 0.02 \\ 
   & 80.00 & 0.33 & \textbf{0.31} & 0.35 & 0.35 & 0.31 \\ 
  False Negatives & 200.00 & \textbf{0.19} & \textbf{0.19} & 0.23 & 0.26 & 0.27 \\ 
   & 1000.00 & \textbf{0.00} & \textbf{0.00} & \textbf{0.00} & 0.01 & 0.03 \\ 
   \hline
\end{tabular}
    \caption{Results from the simulation using the model in Figure 1. Each condition was replicated 200 times. False positives represent concluding that the simulated regressions of zero were concluded as nonzero. False negatives are concluding that either the simulated regression values of 1 or 0.2 are in fact zero. Bolded values represent the smallest error per condition.}
\end{table}

For false positives, the adaptive lasso demonstrated the best
performance, where the performance of maximum likelihood estimation
leveled off at the 0.05 level at a sample size of 1000 as expected. For
false negatives, lasso penalties demonstrated similar results to maximum
likelihood. This was expected given the tendency of the lasso to
under-penalize small coefficients in comparison to the other
regularization methods. The adaptive lasso and SCAD demonstrated
slightly worse results, however, outside of the MCP, each method made
either zero or near zero errors at a sample size of 1000. The poor
performance of the MCP may be in part due to fixing the \(\gamma\)
penalty to 3.7. Varying this parameter may improve the performance of
the method. In summary, the regularization methods demonstrated an
improvement over maximum likelihood, particularly at small samples, for
a model that had a large number of estimated parameters.

@@

Broadly speaking, there is a growing amount of research into the
integration between data mining methods and latent variable models.
Specifically, beyond RegSEM, this has taken the form of item response
theory and regularization (Sun et al. 2016), pairing both structural
equation models with decision trees (Brandmaier et al. 2013),
exploratory psychological network analaysis (e.g. Epskamp, Rhemtulla,
and Borsboom 2016), along with many others. The amount of pairing
between methods that have generally been housed in separate camps will
only increase into the future. This type of research will be facilitated
by the general upsurge in the creation of open source software that
gives users a general framework to test models. This was the motivation
behind creating the \pkg{regsem} package in that users can estimate
models ranging from simple factor analysis models, to latent
longitudinal models with few to many time points, and finally to models
with a large number of latent and observed variables. The use of
regularization allows for the estimation of much larger structural
equation models than before. Specifically, given that latent variables,
as discussed in this article, are more commonly used in the social and
behavioral sciences, sample sizes are generally not large. To estimate
large models with small sample sizes invites increasing amounts of bias
as demonstrated with the simulated data in this paper. Regularization
can be used to reduce the complexity of the model, thus decreasing both
the bias and variance.
